Typed Interactions for NLU in Opal

### 2017-12-13 paper.tex
Added a sentence to the conclusion
Started working on Adrian's suggestions

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{subcaption}

@@

Today, computers are our personal assistants. They are integrated with our lives
and they help everyday people do everyday tasks. Most people who use computers
today never touch a command shell, and even those who do would often rather not.
Graphical interfaces are amazingly powerful, and are often good enough to allow
users to easily interact with the computer, but even those have their drawbacks.
We would really like our personal assistant to respond to commands in human
language:
\begin{itemize}
\item ``Hey Siri! Play some music.''
\item ``Ok Google. Navigate to my hotel.''
\item ``Slackbot, set my status to active.''
\end{itemize}
This is the realm of Natural Language Understanding (NLU)\footnote{We do not
  distinguish between spoken or written language when talking about NLU in
  general.}.

We have seen a huge growth in NLU applications in recent years, and we will
likely see more. Unfortunately, there has been no serious effort made to make
this power available at the language level, so programmers have been forced to
integrate these technologies into their applications by hand. (In fact, machine
learning in general has been relatively neglected by the programming languages
community.) We propose an addition to the Opal language that will begin to close
this gap.

\subsection{An Example}
Consider a simple reminder interface to a calendar or other scheduling
assistant. A user might ask the application to set a reminder containing a
message and a time (for simplicity, we will just allow the user to specify
\ts{later} or \ts{tomorrow}) or she
might ask the application to get any active reminders. Since Opal is embedded
in Typescript, a typical Opal application might express this interface with the
following declarations:\footnote{An overview of Typescript's type system can be
  found at \url{www.typescriptlang.org}.}
\begin{minted}{typescript}
  type Msg = string;
  type Time = "later" | "tomorrow";
  type Set = {
    msg: Msg,
    time: Time
  };
  type Intent =
    | { tag: "Set", data: Set }
    | { tag: "Get", data: {} };
\end{minted}

@@

We will begin by presenting a technical description of our type language in
section \ref{description}. In section \ref{formalism}, we present a formal
specification for the language our type-directed response parsing algorithm.
Section \ref{implementation} covers the implementation of the language and
related libraries. Finally, sections \ref{evaluation} and \ref{future} present
an evaluation of our work, and our plans for future work, respectively.

\section{Language Description} \label{description}
For our initial work, we elected to focus on building a system to support a
single NLU backend---specifically, we are working with an NLU engine called
Wit.\fxnote{[CITE]} When configuring Wit, as with many NLU backends, the main
task is setting up the \emph{entities} that Wit should be looking for. An entity
is a discrete piece of information that the NLU engine extracts from an
utterance. In our example, the messages and the time are both entites, as is the
user's intent.

\subsection{Search Strategies}
When entities are declared, the user must specify a ``search strategy''. In the
case of Wit, there are three main strategies:
\begin{description}
\item[Free Text] Entities that are found via free text matching are just strings
of text from the utterance. In our example, the message is a free text entity.
\item[Keywords] Keywords entities are just words from some pre-defined set.
  Things like \ts{later} and
  \ts{tomorrow} would be captured via keyword searching.
\item[Trait] These are slightly more abstract. Traits capture something that is
  a quality of the utterance as a whole. Something like a user's intent or even
  mood could be a trait.
\end{description}

@@

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
declarations from section \ref{introduction} written in our DSL.

@@

Another benefit of these bespoke declaration forms is simplified syntax. When a
\ts{free-text} entity is declared, there is no need to write ``\ts{= string}'',
because that is implied by the search strategy. Similarly, since traits are
always tagged unions, we can write them in (almost) ML syntax, rather than
writing out the idiomatic Typescript workaround.

Finally, writing specific entity declarations rather than just \ts{type}, makes
it clear that there is more going on than a direct translation into Typescript.
In fact, the DSL parser does lot of validation to guarantee that only reasonable
types can be written down. For example, a \ts{keywords} declaration is required
to be a union, and that union must contain only literals. Even the \ts{alias}
keyword has some validation built in. Some Typescript types would be problematic
if they were allowed---we enforce those conditions in the DSL.

  data Ty = Def String
          | Lit String
          | Rec [(String, Ty)]
          deriving (Show)
\end{minted}
  \caption{The abstract syntax tree for our DSL.}
  \label{fig:ast}
\end{figure}

Figure \ref{fig:ast} shows the AST that is produced when we parse the DSL.
Looking at this, it also becomes clear that the limitations on the types allowed
in a trait declaration are fairly strict. As of writing, there are only three
types allowed within a trait declaration: previously defined entities, literals,
and records (where each element is subject to the same restrictions). These
limitations match up with the capabilities of NLU engines in general. While many
backends provide the ability to parse numbers, dates, and other data, these
types of entities are implementation specific. In section \ref{future}, we
discuss plans to expand the allowed types in a modular way.

@@

We elected to use a Typescript library called \emph{runtypes} to serialize the
AST directly into Typescript; this gives us a data structure to guide response
parsing, and also provides us a way to dynamically type-check the results. We
discuss the response parsing algorithm in more detail in section
\ref{implementation}.

Figure \ref{fig:process} shows process of configuring an application for NLU
interaction. The programmer writes types in our DSL, and uses our Haskell tool
to parse the declarations. The tool serializes three artifacts: a set of
Typescript type declarations, a runtypes datastructure, and a set of JSON
configuration files for Wit. The programmer simply imports the types and
runtypes into their Opal project, uploads the JSON files to Wit, and the system
is ready to go.

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from section \ref{description}. We begin by formalizing the type DSL
as shown in figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{Type DSL}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
    \rangle $$
    \caption{Relevant Opal Types}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{Formal grammars.}
\end{figure}
There are similarities between these definitions and the Haskell ones in figure
\ref{fig:ast}, but these have some slightly different goals. First, we remove
aliases and require that all type definitions are written ``in-line''. Next,
rather than split the definition by allowing tags within $\fcy{T}$, we also
require that entitiy definitions are inlined. This means that the data a user
expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---this
formulation can represent the same types as the actual system. This approach
simply gives up some usability to simplify reasoning.

In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
define a grammar of types, $\tau$. We can think of $\tau$ as representing the
final form of the data that a user hopes to obtain, with no information about
the entities that generated the data. They map almost directly into types that
are used by the underlying Opal system. We add a few types that are noteworthy:
the type {\bf string} is added to account for FreeText entities; and $\langle
{\ell} \rangle$ represents a disjunction of literals, and covers the case of
Keywords entities.

@@

As for the second step, we would like the NLU backend to be set up such that its
responses are \emph{well-formed} with respect to the configuration. Intuitively,
any time we look up some entity tag in the response, we would like to get data
that is compatible with that entity's declaration. We define a type,
\textsf{Resp} of responses, and require that a \ff{lookup} function exists, with
the following type:
$$ \ff{lookup}\ :\ \ff{Tag} \to \ff{Resp} \rightharpoonup \textbf{string} $$
Note that the function is partial; there is no guarantee that the response will
contain the desired data. In the sequel, we will often assume that \ff{lookup}
is defined on tags we care about, but that assumption will be stated explicitly
rather than assumed by default.

@@

We will not write out the full definition of \ff{parseTy}, but note that it is
fairly automatic. If the type is another declaration, we can call back to
\ff{parse}, if it is a literal we simply generate the appropriate string, and if
it is a record, we build the appropriate record and recursively build out its
fields.

@@

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool turns
code in the type DSL into Typescript code and JSON configuration, and ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

Our configuration tool is written in Haskell, and is designed to be simple and
obviously correct. We use the Parsec library to parse the type DSL into the
intermediate representation mentioned above in figure \ref{fig:ast}.

@@

The other difference arises because Opal does not have dependent types. We
cannot completely faithfully transcribe the parse function from above, because
it is not obvious that it can be typed at all. The solution here is to combine
Opal's dynamic typing with runtypes' dynamic checking feature to create a
function that appears to be dependently typed. First, we use dynamic types
within the parse function. While building the response data, we assume, where
necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
runtypes object to check the parsed object, and we cast it to the desired type.
The definition is shown in figure \ref{fig:parse}. The cast is safe, so long as
the user respects the invariant that parse is always called with compatible type
and runtype parameters.

@@

\subsubsection{Confidence-Based Parsing}
Up to this point, we have assumed very little about the data returned in the NLU
response. In practice, however, many NLU engines provide more than just
information about the matched entities. Wit, in particular, provides confidence
levels for each entitiy that it finds. Thus, a value like \ts{"Set"} might be
paired with a confidence level of $70\%$, meaning that Wit is $70\%$ sure that
the user's intent is actually to set a new reminder.

@@

\section{Conclusion} \label{conclusion}
The Opal langauge hopes to be the ideal language for building modern
applications that leverage machine learning techniques. As NLU becomes more and
more popular as a user interface, it makes sense that Opal would provide a clean
and safe interface for natural langauge interaction.

>>>>

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fixme}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{cite}

@@

Today, computers are our personal assistants. They are integrated with our lives
and they help everyday people do everyday tasks. Most people who use computers
today never touch a command shell, and even those who do would often rather not.
Graphical interfaces are amazingly powerful, and are often good enough to allow
users to easily interact with the computer, but even those have their drawbacks.
We would really like our personal assistant to respond to commands in human
language:
\begin{itemize}
\item ``Hey Siri! Play some music.''
\item ``Ok Google. Navigate to my hotel.''
\item ``Slackbot, set my status to active.''
\end{itemize}
This is the realm of Natural Language Understanding (NLU).\footnote{We do not
  distinguish between spoken or written language when talking about NLU in
  general.}

We have seen a huge growth in NLU applications in recent years, and we will
likely see more. Unfortunately, there has been no serious effort made to make
this power available at the language level, so programmers have been forced to
integrate these technologies into their applications by hand. We propose an
addition to the Opal language that will begin to close this gap.

\subsection{An Example}
Consider a simple reminder interface to a calendar or other scheduling
assistant. A user might ask the application to set a reminder containing a
message and a time (for simplicity, we will just allow the user to specify
\ts{later} or \ts{tomorrow}) or she might ask the application to get any active
reminders. Since Opal is embedded in TypeScript,\cite{typescript} a typical Opal
application might express this interface with the following declarations:
\begin{minted}{typescript}
  type Msg = string;
  type Time = "later" | "tomorrow";
  type Set = {
    msg: Msg,
    time: Time
  };
  type Intent =
    | { tag: "Set", data: Set }
    | { tag: "Get", data: {} };
\end{minted}

@@

We will begin by presenting a technical description of our type language in
Section \ref{description}. In Section \ref{formalism}, we present a formal
specification for the language our type-directed response parsing algorithm.
Section \ref{implementation} covers the implementation of the language and
related libraries. Finally, Sections \ref{evaluation} and \ref{future} present
an evaluation of our work, and our plans for future work, respectively.

\section{Language Description} \label{description}
For our initial work, we elected to focus on building a system to support a
single NLU backend---specifically, we are working with an NLU engine called
Wit.\cite{wit} When configuring Wit, as with many NLU backends, the main task is
setting up the \emph{entities} that Wit should be looking for. An entity is a
discrete piece of information that the NLU engine extracts from an utterance. In
our example, the messages and the time are both entites, as is the user's
intent.

\subsection{Search Strategies}
When entities are declared, the user must specify a ``search strategy.'' In the
case of Wit, there are three main strategies:
\begin{description}
\item[Free Text] Entities that are found via free text matching are just strings
of text from the utterance. In our example, the message is a free text entity.
\item[Keywords] Keywords entities are just words from some pre-defined set.
  Things like \ts{later} and
  \ts{tomorrow} would be captured via keyword searching.
\item[Trait] These are slightly more abstract. Traits capture something that is
  a quality of the utterance as a whole. Something like a user's intent or even
  mood could be a trait.
\end{description}

@@

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
declarations from Section \ref{introduction} written in our DSL.

@@

Another benefit of these bespoke declaration forms is simplified syntax. When a
\ts{free-text} entity is declared, there is no need to write ``\ts{= string}'',
because that is implied by the search strategy. Similarly, since traits are
always tagged unions, we can write them in (almost) ML syntax, rather than
writing out the idiomatic TypeScript workaround.

Finally, writing specific entity declarations rather than just \ts{type}, makes
it clear that there is more going on than a direct translation into TypeScript.
In fact, the DSL parser does lot of validation to guarantee that only reasonable
types can be written down. For example, a \ts{keywords} declaration is required
to be a union, and that union must contain only literals. Even the \ts{alias}
keyword has some validation built in. Some TypeScript types would be problematic
if they were allowed---we enforce those conditions in the DSL.

\begin{figure}
  \centering
  \begin{align*}
    \mathbb{D} ::=&\ \text{FreeText}(\etag{t}) \\
    |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
    |&\ \text{Trait}(\etag{t},\ \{s: \mathbb{T}\}) \\
    |&\ \text{Alias}(\etag{t},\ \mathbb{T}) \\
    \\
    \mathbb{T} ::=&\ \text{Def}(\etag{t}) \mid \ell \mid \{s: \mathbb{T}\}
  \end{align*}
  $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
  \caption{The abstract syntax tree for our DSL.}
  \label{fig:ast}
\end{figure}

Figure \ref{fig:ast} shows the AST that is produced when we parse the DSL.
Looking at this, it also becomes clear that the limitations on the types allowed
in a trait declaration are fairly strict. As of writing, there are only three
types allowed within a trait declaration: previously defined entities, literals,
and records (where each element is subject to the same restrictions). These
limitations match up with the capabilities of NLU engines in general. While many
backends provide the ability to parse numbers, dates, and other data, these
types of entities are implementation specific. In Section \ref{future}, we
discuss plans to expand the allowed types in a modular way.

@@

We elected to use a TypeScript library called \emph{runtypes} to serialize the
AST directly into TypeScript; this gives us a data structure to guide response
parsing, and also provides us a way to dynamically type-check the results. We
discuss the response parsing algorithm in more detail in Section
\ref{implementation}.

Figure \ref{fig:process} shows process of configuring an application for NLU
interaction. The programmer writes types in our DSL, and uses our Haskell tool
to parse the declarations. The tool serializes three artifacts: a set of
TypeScript type declarations, a runtypes datastructure, and a set of JSON
configuration files for Wit. The programmer simply imports the types and
runtypes into their Opal project, uploads the JSON files to Wit, and the system
is ready to go.

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results from Section \ref{description}. We begin by formalizing the type DSL
as shown in Figure \ref{fig:grammar_a}.
\begin{figure}
  \centering
  \begin{subfigure}{1\linewidth}
    \begin{align*}
      \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
      |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
      \\
      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
    \end{align*}
    $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
    \caption{Type DSL}
    \label{fig:grammar_a}
  \end{subfigure}\vspace{1cm}
  \begin{subfigure}{1\linewidth}
    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
    \rangle $$
    \caption{Relevant Opal Types}
    \label{fig:grammar_b}
  \end{subfigure}
  \caption{The core language.}
\end{figure}
There are similarities between these definitions and the ones in Figure
\ref{fig:ast}, but the core language has some slightly different goals. First,
we remove aliases and require that all type definitions are written ``in-line.''
Next, rather than split the definition by allowing tags within $\fcy{T}$, we
also require that entity definitions are inlined. This means that the data a
user expects from the NLU backend can be represented by a single tree of type
$\fcy{D}$. Importantly, none of these simplifications are fundamental---this
formulation can represent the same types as the actual system. This approach
simply gives up some usability to simplify reasoning.

In addition to a grammar for the core language itself, in Figure
\ref{fig:grammar_b} we define a grammar of types, $\tau$. We can think of $\tau$
as representing the final form of the data that a user hopes to obtain, with no
information about the entities that generated the data. They map almost directly
into types that are used by the underlying Opal system. We add a few types that
are noteworthy: the type {\bf string} is added to account for FreeText entities;
and $\langle {\ell} \rangle$ represents a disjunction of literals, and covers
the case of Keywords entities.

@@

As for the second step, we would like the NLU backend to be set up such that its
responses are \emph{well-formed} with respect to the configuration. Intuitively,
any time we look up some entity tag in the response, we would like to get data
that is compatible with that entity's declaration. We define a type,
\textsf{Resp} of responses, and require that a \ff{lookup} function exists, with
the following type:
$$ \ff{lookup}\ :\ \ff{Tag} \to \ff{Resp} \rightharpoonup \textbf{string} $$
Note that the function is partial; there is no guarantee that the response will
contain the desired data. In the following development, we will often assume
that \ff{lookup} is defined on tags we care about, but that assumption will be
stated explicitly rather than assumed by default.

@@

We will not write out the full definition of \ff{parseTy}, but note that it is
fairly automatic. If the type is another declaration, we can call back to
\ff{parse}: if it is a literal we simply generate the appropriate string; and if
it is a record, we build the appropriate record and recursively build out its
fields.

@@

\subsection{Configuration Tool}
When a user wants to configure an Opal application to interact with an NLU
library like Wit, they begin by using our configuration tool. The tool turns
code in the type DSL into TypeScript code and JSON configuration, and ensures
that the NLU engine responses will agree with the type definitions in the user's
application.

Our configuration tool is written in Haskell, and is designed to be simple and
obviously correct. We use the Parsec library to parse the type DSL into the
intermediate representation mentioned in Figure \ref{fig:hs_ast}.

  data Ty = Def String
          | Lit String
          | Rec [(String, Ty)]
          deriving (Show)
\end{minted}
  \caption{The Haskell code for our AST.}
  \label{fig:hs_ast}
\end{figure}

@@

The other difference arises because Opal does not have dependent types. We
cannot completely faithfully transcribe the parse function from above, because
it is not obvious that it can be typed at all. The solution here is to combine
Opal's dynamic typing with runtypes' dynamic checking feature to create a
function that appears to be dependently typed. First, we use dynamic types
within the parse function. While building the response data, we assume, where
necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
runtypes object to check the parsed object, and we cast it to the desired type.
The definition is shown in Figure \ref{fig:parse}. The cast is safe, so long as
the user respects the invariant that parse is always called with compatible type
and runtype parameters.

@@

\subsubsection{Confidence-Based Parsing}
Up to this point, we have assumed very little about the data returned in the NLU
response. In practice, however, many NLU engines provide more than just
information about the matched entities. Wit, in particular, provides confidence
levels for each entity that it finds. Thus, a value like \ts{"Set"} might be
paired with a confidence level of $70\%$, meaning that Wit is $70\%$ sure that
the user's intent is actually to set a new reminder.

@@

\section{Conclusion} \label{conclusion}
The Opal langauge hopes to be the ideal language for building modern
applications that leverage machine learning techniques. As NLU becomes more and
more popular as a user interface, it makes sense that Opal would provide a clean
and safe interface for natural langauge interaction. This work contributes a
real, working tool in the Opal ecosystem that provides a first attempt at such
an interface.

\bibliography{paper}{}
\bibliographystyle{plain}
